%! Author = Martin Vandenbussche

\documentclass[a4paper,11pt]{scrreprt}
\input{header.tex}

\title{Master Thesis - A new syntax for the Oz programming language}
\subject{Under the supervision of Prof. Peter Van Roy}
\author{Martin Vandenbussche - 02441500}
\date{Academic year 2020--2021}

\begin{document}

\maketitle

\input{abstract.tex}

\begin{comment}
Section 1.1 Contexte et problème
Section 1.2: Inspirations
1.2.1: Scala
1.2.2 Ozma
1.2.3 Jean-Pacifique
Section 1.3 Contributions
1.3.1 Adaptation du travail Jean-Pacifique
1.3.2 NewOz compiler
1.3.3 Community feedback
Section 1.4 Final conclusion on new syntax

chap 2 -> Design principles + my april syntax

chap 3 -> nozc compiler

chap 4 -> community feedback
Comprend  une conclusion qui donne une syntaxe newOz finale avec ton expérience y compris feedback de la communauté

chap 5 -> résumé du processus complet

Appendices : grammar, translations exmaples of Oz to NewOz, tutorial/doc from GitHub
\end{comment}

\tableofcontents

%\medskip

\chapter{Goal of the project and previous works}\label{ch:1}
\input{chapter1.tex}

\chapter{Design principles of the new syntax}\label{ch:2}
\input{chapter2.tex}

\chapter{The NewOz Compiler : nozc}\label{ch:3}
\input{chapter3.tex}

\chapter{The feedback from the community}\label{ch:4}
\input{chapter4.tex}

\chapter{Conclusion}\label{ch:5}
\input{conclusion.tex}

\begin{appendices}
\input{appendices.tex}
\end{appendices}

\printbibliography



\begin{comment}
Promoters:
Peter Van Roy
Other people :
Laurent Nicolas (INGI)
Description:
Oz is a well-factored language that supports many programming paradigms and is successfully being used in courses and MOOCs.
It has also influenced the development of other languages such as Scala and Go.
The internal structure of Oz has stood the test of time.
But the syntax has not: there are many idioms used today that Oz syntax supports poorly.
For example, big-data style computations often use a composition of functional objects, with a syntax something like DB.select_list(f1).map(f2).sort(f3).reduce(f4,u), with a chain of operations such that each operation takes a functional object and returns a new functional object.
Oz syntax does not support this well.

This master's project continues the work of an earlier master's project on improving Oz syntax.
The ultimate goal is to change the Oz syntax in the official release of Mozart 2, which is a major result.
That is why this work is spread over several years, in several master's projects.
The goal is to make a completely new syntax of Oz, taking advantage of the best parts of Scala, Python, Go, Clojure, and other modern languages.
The new syntax should be simple, well-factored, and support sophisticated programming idioms, while still keeping the clean Oz kernel language semantics.
You will study important programming idioms in many languages and design and implement a new parser front end for the Mozart 2 system.

References
[1] Peter Van Roy and Seif Haridi.  Concepts, Techniques, and Models of Computer Programming.  MIT Press, 2004.
[2] Peter Van Roy, Seif Haridi, Christian Schulte, and Gert Smolka.  A History of the Oz Multiparadigm Language.  The Fourth ACM SIGPLAN History of Programming Languages Conference, 2020.
[3] The Mozart Programming System, http://www.mozart2.org
[4] Welcome to Python.org, http://www.python.org
[5] The Scala Programming Language, http://www.scala-lang.org
[6] Erlang Programming Language, https://www.erlang.org
[7] The Go Programming Language, https://golang.org
[8] Clojure, https://clojure.org
[9] François Fonteyn, Comprehensions in Mozart, Master's thesis, ICTEAM Institute, Université catholique de Louvain, June 2014.
\end{comment}


\end{document}